The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-11 paper/ms.tex
Change to typewriter for orbits, fix softening lengths

\newcommand{\thincolor}{pink}
\newcommand{\thickcolor}{brown}
\newcommand{\halocolor}{blue}

@@

\begin{deluxetable*}{cCCCCCCCCC}
\tablecaption{Description and names of the three orbits considered in this
work. For the last four columns: $z_{\text{max}}$ is the maximum height of the
orbit, $\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$ is the magnitude of
the radial excursions of the orbit, and $\kappa$ and $\nu$ are the radial/epicyclic and vertical frequencies of the orbit. In the epicyclic approximation, $A_z =
z_{\text{max}}$ and $A_R = \frac{1}{2}(R_{\text{max}} - R_{\text{min}})$.
\label{tab:orbits}}
\tablehead{\colhead{name} & \colhead{initial position} & \colhead{initial
velocity} & \colhead{$J_R$} &
\colhead{$J_{\phi}$} & \colhead{$J_z$} & \colhead{$z_{\text{max}}$} &
\colhead{$\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$} & \colhead{$\kappa$} & \colhead{$\nu$} \\ \colhead{ } &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{km\,s}^{-1}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{Myr}^{-1}$} & \colhead{$\mathrm{Myr}^{-1}$}}
\startdata 
thin-disk & (8, 0, 0) & (0, -190, 10) & 40.3 & -1520 & 0.69 & 0.12 & 1.29 & 0.049 & 0.093 \\
thick-disk & (8, 0, 0) & (0, -190, 50) & 32.5 & -1520 & 23.0 & 0.85 & 1.19 & 0.048 & 0.061 \\ 
halo & (8, 0, 0) & (0, -190, 190) & 32.8 & -1520 & 529.1 & 6.16 & 2.34 & 0.033 & 0.025
\enddata
\end{deluxetable*}

@@

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the computed actions with an
error in the Galactocentric coordinate system. We consider here the thick-disk
orbit, which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520,
7.0)\,\actunit$ and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We
integrate the orbit according to the procedure laid out in
Section~\ref{ssec:action_comp}, and which we plot in
Appendix~\ref{app:orbits}. Then, we subtract $100\,\pc$ from the $z$ value
(upper panels) or the $x$ value (lower panels) of each position in the orbit,
corresponding to an erroneous observer assuming a midplane (upper) or solar
radius (lower) that is off by $100\,\pc$. We then allow an immortal observer
to observe the orbit over $1\,\Gyr$ and perform the same orbit integration
procedure at each timestep, and report the values of the actions. The
computation of $J_{\phi}$ is pristine to errors in $z$, with only numerical
artifacts remaining. Only small errors are induced in $J_R$, with the middle
$90\%$ of values over the $\Gyr$ being within $\sim8\%$ of the true $J_R$. As
expected, large errors are induced in $J_z$ with a $100\,\pc$ offset in $z$,
with the middle $90\%$ of values being within $\sim43\%$ of the true $J_z$.
The $x$ offset induces uncertainties in $J_R$, $J_{\phi}$, and $J_z$ of
$\sim21\%$, $\sim3\%$, and $\sim3\%$.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the computed values of $J_z$ at different orbital phases for the thick-disk (top
panel) and thin-disk (bottom panel) orbits assuming a $z$ offset of
$100\,\pc$. One can see that if the observed $z$ values have a bias (from, e.g.,
an incorrectly computed midplane), then the induced error distribution in
$J_z$ is decidedly non-Gaussian. Therefore, any sort of error propagation must
take this into account. Intuition for the shape of each panel is given in the
text. We also plot one half the 95\uth{} percentile minus the 5\uth{}
percentile of each distribution as a horizontal arrow anchored on the true
$J_z$ value. We call this $\Delta J_z$ and will use it (along with the
similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to empirically describe
the error distribution. We see that $\Delta J_z$ roughly corresponds to the
distance from the true $J_z$ value to one of the modes of the distribution of
computed $J_z$ values. Similar plots for $J_R$ induced by a $z$ offset and
$J_R$ and $J_{\phi}$ induced by an $x$ offset are given in
Appendix~\ref{app:hist}.}
\label{fig:Jz_hist}
\end{figure}

We now quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the actions calculated from the observed position and velocity of a star. We consider
three orbits in the model potential described in
Section~\ref{ssec:action_comp} that are typical of stars in the thin
disk, thick disk, and halo. We summarize their initial
positions in phase space, the actions computed by integrating their orbits
in the correct potential, and other properties in Table~\ref{tab:orbits}. Each orbit, integrated
without systematic coordinate errors, is plotted in Appendix~\ref{app:orbits}.
We will refer to these orbits by their names (thin-disk, thick-disk, halo)
henceforth.

We begin with the thick-disk orbit. Consider an observer who can
measure the thick-disk star's phase-space position at many different times
(and hence different orbital phases), but does so using a
coordinate system in which the midplane is systematically offset in height by
$100\,\pc$ from its actual location. To model this we subtract the vector $(0, 0,
100)\,\pc$ from each position in the orbit. This corresponds to an
observer physically located at, e.g., the position $(8, 0, 0)\,\kpc$ in the
coordinate system of the true potential, but erroneously thinking they are
located at $(8, 0, 0.1)\,\kpc$.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the values of
$J_z$ computed at different orbital phases for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}). The true value is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the null value. In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is~$\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit (equivalent to $A_z$ in the epicyclic approximation, see
Section~\ref{ssec:epi_action}).

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The left, center, and right panels show the result for $J_R$,
$J_{\phi}$, and $J_z$, respectively. The upper panels consider an offset in
$z$ and the lower panels consider an offset in $x$ (equivalently, an offset in
the Solar radius). In some panels we also plot as dashed lines the epicyclic
prediction of the induced action error (Equation~\eqref{eq:Ji_err_mosttime}).
In the epicyclic approximation, a $z$ offset only induces an error in $J_z$
--- for all three orbits the epicyclic approximation is a good description of
the $J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The
error in $J_R$ is somewhat well-described for the thin-disk and thick-disk
orbits, and a poor description for the halo orbit. For $J_{\phi}$, the
epicyclic approximation is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the thick-disk orbit (top panel), where the offset in $z$ is
much less than $z_{\text{max}}$ of the orbit. The peaks in the distribution
correspond to the turning points of the orbit (or points of maximum vertical
amplitude), where $v_z \sim 0$ and where the star is on most of it orbit. This
is why the distribution, which is calculated at evenly spaced time intervals, peaks at these values. For the thin-disk orbit
(bottom panel), the offset in $z$ is comparable to $z_{\text{max}}$. Now,
there will be some points in the orbit where $v_z = 0$ and $z=0$ (in the
erroneous coordinate system). At these points, the computed $J_z$
will vanish. The asymmetry and systematic offset then comes about because of
the constraint that $J_z \geq 0$.\footnote{This argument is similar to ones given in
cosmology for why gravity produces non-Gaussianity in the density field, since
the density cannot become negative but it can grow arbitrarily large.}

@@

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report $\Delta
J_i/J_i$ for the three different fiducial orbits in Table~\ref{tab:orbits}.
The thick-disk (\thickcolor) orbit is the one from
Figure~\ref{fig:one_orbit_wrong_ref}, but we also now consider the effect on
the action determined for the thin-disk (\thincolor) and halo (\halocolor)
orbits.

@@

In the upper middle panel of Figure~\ref{fig:many_orbit_wrong_ref}, there is
essentially no spread in the determination of $J_{\phi}$. This is expected
since $J_{\phi}$ is independent of $z$ and is thus unaffected by offsets in
$z$, as discussed earlier. Indeed, the result we found in
Figure~\ref{fig:one_orbit_wrong_ref} for the thick-disk orbit holds for all
orbit types. This is also a demonstration of the robustness of the
integration and action calculation methods we use.

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For the thin-disk (thick-disk) orbit, a systematic offset of $15\,\pc$ ($120\,\pc$) in the $z$-coordinate results in a $25\%$ deviation in $J_z$.
The halo orbit is relatively resistant to errors in the midplane, with
only $\sim15\%$ error in $J_z$ out to an offset of $500\,\pc$.

@@

The epicyclic approximation is a good predictor of $\Delta J_z$, even for halo
orbits. It performs similarly for $\Delta J_R$, now underpredicting for halo
orbits and slightly overpredicting for thin-disk orbits. Note that for the
particular orbits we chose, the thin-disk orbit has slightly larger $A_R$ than
the thick-disk orbit, and so we expect the epicyclic approximation to perform
slightly worse. The epicyclic approximation underpredicts $\Delta J_{\phi}$
for all orbits.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(0, -190, v_z)\,\kms$, where we vary $v_z$. We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green
respectively). As before, the error ($\Delta J_z$) is one half the 95\uth{}
minus 5\uth{} percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for thin-disk like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$.\footnote{The thin-disk, thick-disk, and halo orbits correspond to setting $v_z=10$, $50$, and $190$, respectively (see Table~\ref{tab:orbits}).} For a thin disk-like orbit ($J_z\sim0.5\,\actunit$), even a
$10\,\pc$ offset in $z$ is enough to induce a $\sim25\%$ error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

@@

\begin{deluxetable*}{cccccc}[htb!]
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work. For comparison, we also give the median
softening lengths for the simulated galaxies, computed for cold gas ($T<1000\,\text{K}$) and stars with $\abs{R-R_0} < 0.5\,\kpc$ and $\abs{z} < 1\,\kpc$. We have assumed that $R_0=8.2\,\kpc$.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{\makecell{cold\tablenotemark{a} gas disk \\ scale height}} &
\colhead{\makecell{thin disk \\ scale height}} & \colhead{\makecell{thick disk \\ scale height}} & \colhead{\makecell{cold\tablenotemark{a} gas \\
softening length}} & \colhead{\makecell{stellar \\ softening length}} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
Milky Way\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

These simulations contain dark matter particles of mass $\sim35,000\,\Msun$,
gas particles of mass $\sim7000$ to $20,000\,\Msun$, and star particles of
mass $\sim5000$ to $7000\,\Msun$, with the lower end coming from
stellar evolution \citep{2018arXiv180610564S}. Softening lengths for dark
matter and star particles are fixed at $112\,\pc$ and $11.2\,\pc$,
respectively.\footnote{This is $2.8$ times the often-quoted
Plummer-equivalent.} The gas softening length is adaptive, but at $\z=0$ the
median softening length for cold ($T < 10^4\,\text{K}$) gas particles around
roughly solar positions (with galactocentric cylindrical radii within
$0.5\,\kpc$ of $8.2\,\kpc$ and $\abs{z}<1\,\kpc$) is $12.4$, $12.5$, and
$10.3\,\pc$ for \mi{}, \mf{}, and \mm{}, respectively. These values are
summarized in Table~\ref{tab:scale_height}, along with measurements of the
stellar and gas disk scale heights.

The softening lengths used in the simulations can affect the ability to
resolve the very thinnest planar structures, which in turn can affect how much
the density-based midplane varies as a function of azimuth. The Milky Way's
dense, star-forming gas disk is thought to have a scale height of about
$40\,\pc$, $\sim3\textup{--}4$ times the cold gas softening length
\citep{2019ApJ...871..145A}. The thin stellar disk has a scale height of about
$300\,\pc$, $\sim30$ times the stellar softening length
\citep{2008ApJ...673..864J}. We therefore expect that resolution effects may
still be affecting the scale heights of these components in the simulations,
especially the cold gas. Indeed, the stellar scale heights of the simulated
galaxies are equal to or larger than the Milky Way's while the gas scale
heights are significantly larger (although the proper basis comparison is less
clear in the case of the gas; the quoted value for the Milky Way comes from
studies of high-mass star-forming regions). As noted above, the midplanes
defined by gas and stars can be tilted with respect to one another as well,
precluding extending the precision of the gas midplane definition to the
stellar component.

@@

The vertical frequency of the thin-disk and thick-disk orbits are $\sim
0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$, respectively (Table~\ref{tab:orbits}). By dimensional
analysis, and assuming a vertical velocity variation of $5\textup{--}10\,\kms$, we
therefore expect the midplane offsets to be $\sim 57\textup{--}170\,\pc$. We
stress that this is a rough calculation.

>>>>

\newcommand{\thin}{\texttt{thin-disk}}
\newcommand{\thick}{\texttt{thick-disk}}
\newcommand{\halo}{\texttt{halo}}
\newcommand{\thincolor}{pink}
\newcommand{\thickcolor}{brown}
\newcommand{\halocolor}{blue}

@@

\begin{deluxetable*}{cCCCCCCCCC}
\tablecaption{Description and names of the three orbits considered in this
work. For the last four columns: $z_{\text{max}}$ is the maximum height of the
orbit, $\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$ is the magnitude of
the radial excursions of the orbit, and $\kappa$ and $\nu$ are the radial/epicyclic and vertical frequencies of the orbit. In the epicyclic approximation, $A_z =
z_{\text{max}}$ and $A_R = \frac{1}{2}(R_{\text{max}} - R_{\text{min}})$.
\label{tab:orbits}}
\tablehead{\colhead{name} & \colhead{initial position} & \colhead{initial
velocity} & \colhead{$J_R$} &
\colhead{$J_{\phi}$} & \colhead{$J_z$} & \colhead{$z_{\text{max}}$} &
\colhead{$\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$} & \colhead{$\kappa$} & \colhead{$\nu$} \\ \colhead{ } &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{km\,s}^{-1}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{Myr}^{-1}$} & \colhead{$\mathrm{Myr}^{-1}$}}
\startdata 
\thin{} & (8, 0, 0) & (0, -190, 10) & 40.3 & -1520 & 0.69 & 0.12 & 1.29 & 0.049 & 0.093 \\
\thick{} & (8, 0, 0) & (0, -190, 50) & 32.5 & -1520 & 23.0 & 0.85 & 1.19 & 0.048 & 0.061 \\ 
\halo{} & (8, 0, 0) & (0, -190, 190) & 32.8 & -1520 & 529.1 & 6.16 & 2.34 & 0.033 & 0.025
\enddata
\end{deluxetable*}

@@

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the computed actions with an
error in the Galactocentric coordinate system. We consider here \thick{},
which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520,
7.0)\,\actunit$ and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We
integrate the orbit according to the procedure laid out in
Section~\ref{ssec:action_comp}, and which we plot in
Appendix~\ref{app:orbits}. Then, we subtract $100\,\pc$ from the $z$ value
(upper panels) or the $x$ value (lower panels) of each position in the orbit,
corresponding to an erroneous observer assuming a midplane (upper) or solar
radius (lower) that is off by $100\,\pc$. We then allow an immortal observer
to observe the orbit over $1\,\Gyr$ and perform the same orbit integration
procedure at each timestep, and report the values of the actions. The
computation of $J_{\phi}$ is pristine to errors in $z$, with only numerical
artifacts remaining. Only small errors are induced in $J_R$, with the middle
$90\%$ of values over the $\Gyr$ being within $\sim8\%$ of the true $J_R$. As
expected, large errors are induced in $J_z$ with a $100\,\pc$ offset in $z$,
with the middle $90\%$ of values being within $\sim43\%$ of the true $J_z$.
The $x$ offset induces uncertainties in $J_R$, $J_{\phi}$, and $J_z$ of
$\sim21\%$, $\sim3\%$, and $\sim3\%$.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the computed values of $J_z$ at different orbital phases for \thick{} (top
panel) and \thin{} (bottom panel) orbits assuming a $z$ offset of
$100\,\pc$. One can see that if the observed $z$ values have a bias (from, e.g.,
an incorrectly computed midplane), then the induced error distribution in
$J_z$ is decidedly non-Gaussian. Therefore, any sort of error propagation must
take this into account. Intuition for the shape of each panel is given in the
text. We also plot one half the 95\uth{} percentile minus the 5\uth{}
percentile of each distribution as a horizontal arrow anchored on the true
$J_z$ value. We call this $\Delta J_z$ and will use it (along with the
similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to empirically describe
the error distribution. We see that $\Delta J_z$ roughly corresponds to the
distance from the true $J_z$ value to one of the modes of the distribution of
computed $J_z$ values. Similar plots for $J_R$ induced by a $z$ offset and
$J_R$ and $J_{\phi}$ induced by an $x$ offset are given in
Appendix~\ref{app:hist}.}
\label{fig:Jz_hist}
\end{figure}

We now quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the actions calculated from the observed position and velocity of a star. We consider
three orbits in the model potential described in
Section~\ref{ssec:action_comp} that are typical of stars in the thin
disk, thick disk, and halo. We summarize their initial
positions in phase space, the actions computed by integrating their orbits
in the correct potential, and other properties in Table~\ref{tab:orbits}. Each orbit, integrated
without systematic coordinate errors, is plotted in Appendix~\ref{app:orbits}.
We will refer to these orbits by their names (\thin{}, \thick{}, \halo{})
henceforth.

We begin with \thick{}. Consider an observer who can
measure the orbit's phase-space position at many different times
(and hence different orbital phases), but does so using a
coordinate system in which the midplane is systematically offset in height by
$100\,\pc$ from its actual location. To model this we subtract the vector $(0, 0,
100)\,\pc$ from each position in the orbit. This corresponds to an
observer physically located at, e.g., the position $(8, 0, 0)\,\kpc$ in the
coordinate system of the true potential, but erroneously thinking they are
located at $(8, 0, 0.1)\,\kpc$.

@@

In Figure~\ref{fig:Jz_hist}, we plot a histogram of the values of
$J_z$ computed at different orbital phases for \thick{} (top panel) and \thin{} (bottom panel),
assuming a $z$ offset of $100\,\pc$ (as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}). The true value is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the null value. In the case of \thin{}
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is~$\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit (equivalent to $A_z$ in the epicyclic approximation, see
Section~\ref{ssec:epi_action}).

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for
\thin{}, \thick{}, and \halo{} orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The left, center, and right panels show the result for $J_R$,
$J_{\phi}$, and $J_z$, respectively. The upper panels consider an offset in
$z$ and the lower panels consider an offset in $x$ (equivalently, an offset in
the Solar radius). In some panels we also plot as dashed lines the epicyclic
prediction of the induced action error (Equation~\eqref{eq:Ji_err_mosttime}).
In the epicyclic approximation, a $z$ offset only induces an error in $J_z$
--- for all three orbits the epicyclic approximation is a good description of
the $J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The
error in $J_R$ is somewhat well-described for \thin{} and \thick{}, 
and a poor description for \halo{}. For $J_{\phi}$, the
epicyclic approximation is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the \thick{} (top panel), where the offset in $z$ is
much less than $z_{\text{max}}$ of the orbit. The peaks in the distribution
correspond to the turning points of the orbit (or points of maximum vertical
amplitude), where $v_z \sim 0$ and where the star is on most of it orbit. This
is why the distribution, which is calculated at evenly spaced time intervals, peaks at these values. For the \thin{}
(bottom panel), the offset in $z$ is comparable to $z_{\text{max}}$. Now,
there will be some points in the orbit where $v_z = 0$ and $z=0$ (in the
erroneous coordinate system). At these points, the computed $J_z$
will vanish. The asymmetry and systematic offset then comes about because of
the constraint that $J_z \geq 0$.\footnote{This argument is similar to ones given in
cosmology for why gravity produces non-Gaussianity in the density field, since
the density cannot become negative but it can grow arbitrarily large.}

@@

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report $\Delta
J_i/J_i$ for the three different fiducial orbits in Table~\ref{tab:orbits}.
Figure~\ref{fig:one_orbit_wrong_ref} considered \thick{} (\thickcolor), but we also now consider the effect on
the action determined for \thin{} (\thincolor) and \halo{} (\halocolor).

@@

In the upper middle panel of Figure~\ref{fig:many_orbit_wrong_ref}, there is
essentially no spread in the determination of $J_{\phi}$. This is expected
since $J_{\phi}$ is independent of $z$ and is thus unaffected by offsets in
$z$, as discussed earlier. Indeed, the result we found in
Figure~\ref{fig:one_orbit_wrong_ref} for \thick{} holds for all
orbit types. This is also a demonstration of the robustness of the
integration and action calculation methods we use.

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For \thin{} (\thick{}), a systematic offset of $15\,\pc$ ($120\,\pc$) in the $z$-coordinate results in a $25\%$ deviation in $J_z$.
We find that \halo{} is relatively resistant to errors in the midplane, with
only $\sim15\%$ error in $J_z$ out to an offset of $500\,\pc$.

@@

The epicyclic approximation is a good predictor of $\Delta J_z$, even for \halo{}.
It performs similarly for $\Delta J_R$, now underpredicting for \halo{}
and slightly overpredicting for \thin{}. Note that for the
particular orbits we chose, \thin{} has slightly larger $A_R$ than
\thick{}, and so we actually expect the epicyclic approximation to perform
slightly worse for \thin{} in this case. The epicyclic approximation underpredicts $\Delta J_{\phi}$
for all orbits.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(0, -190, v_z)\,\kms$, where we vary $v_z$. We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green
respectively). As before, the error ($\Delta J_z$) is one half the 95\uth{}
minus 5\uth{} percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for \thin{} like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$.\footnote{One can recover \thin{}, \thick{}, and \halo{} by setting $v_z=10$, $50$, and $190$, respectively (see Table~\ref{tab:orbits}).} For a \thin{}-like orbit ($J_z\sim0.5\,\actunit$), even a
$10\,\pc$ offset in $z$ is enough to induce a $\sim25\%$ error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

@@

\begin{deluxetable*}{cccccc}[htb!]
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work. For comparison, we also give the median
softening lengths for the simulated galaxies, computed for cold gas ($T<1000\,\text{K}$) and stars with $\abs{R-R_0} < 0.5\,\kpc$ and $\abs{z} < 1\,\kpc$. We have assumed that $R_0=8.2\,\kpc$.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{\makecell{cold\tablenotemark{a} gas disk \\ scale height}} &
\colhead{\makecell{thin disk \\ scale height}} & \colhead{\makecell{thick disk \\ scale height}} & \colhead{\makecell{cold\tablenotemark{a} gas \\
softening length}} & \colhead{\makecell{stellar \\ softening length}} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
Milky Way\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 53.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 57.2 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 & 60.1 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<100\,\text{K}$}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

These simulations contain dark matter particles of mass $\sim35,000\,\Msun$,
gas particles of mass $\sim7000$ to $20,000\,\Msun$, and star particles of
mass $\sim5000$ to $7000\,\Msun$, with the lower end coming from
stellar evolution \citep{2018arXiv180610564S}. Softening lengths for dark
matter and star particles are fixed at $112\,\pc$ and $11.2\,\pc$,
respectively.\footnote{This is $2.8$ times the often-quoted
Plummer-equivalent.} The gas softening length is adaptive, but at $\z=0$ the
median softening length for cold ($T < 100\,\text{K}$) gas particles around
roughly solar positions (with galactocentric cylindrical radii within
$0.5\,\kpc$ of $8.2\,\kpc$ and $\abs{z}<1\,\kpc$) is $53.4$, $57.2$, and
$60.1\,\pc$ for \mi{}, \mf{}, and \mm{}, respectively. These values are
summarized in Table~\ref{tab:scale_height}, along with measurements of the
stellar and gas disk scale heights.

The softening lengths used in the simulations can affect the ability to
resolve the very thinnest planar structures, which in turn can affect how much
the density-based midplane varies as a function of azimuth. The Milky Way's
dense, star-forming gas disk is thought to have a scale height of about
$40\,\pc$, on the order of the cold gas softening length
\citep{2019ApJ...871..145A}. The thin stellar disk has a scale height of about
$300\,\pc$, $\sim30$ times the stellar softening length
\citep{2008ApJ...673..864J}. We therefore expect that resolution effects are
still affecting the scale heights of these components in the simulations,
especially the cold gas. Indeed, the stellar scale heights of the simulated
galaxies are equal to or larger than the Milky Way's while the gas scale
heights are significantly larger (although the proper basis comparison is less
clear in the case of the gas; the quoted value for the Milky Way comes from
studies of high-mass star-forming regions). As noted above, the midplanes
defined by gas and stars can be tilted with respect to one another as well,
precluding extending the precision of the gas midplane definition to the
stellar component.

@@

The vertical frequency of \thin{} and \thick{} are $\sim
0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$, respectively (Table~\ref{tab:orbits}). By dimensional
analysis, and assuming a vertical velocity variation of $5\textup{--}10\,\kms$, we
therefore expect the midplane offsets to be $\sim 57\textup{--}170\,\pc$. We
stress that this is a rough calculation.
